Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-07 paper/manuscript.tex
more fixes

However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin. One way to look at this is to examine how low $\lambda$ must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep. If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T){\frac}{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.  So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation! This is not very much like the desired final state of no interactions between the particle and its environment. We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$. Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble}

@@

The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
$U_0(\vec{q})$ and $U_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
\begin{eqnarray}
\Delta f_{01} & = & \ln \expect{e^{-(U_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
\end{eqnarray}\label{eqn.zwanzig}
and the average is over all samples from the simulation perfomed with $u_0$. In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $A/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.   This almost requires no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.  The analysis can be written in a line of code.  We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{Jarzynski:PRL:1997,Jarzynski:APPB:1998,Crooks:PRE:2000}, though we do not go into all of the details of nonequilibrum transformations here, and refer the reader to more advanced treatments~\cite{Maragakis:JCP:2008,Oberhofer:JPCB:2005,Procacci:JCP:2015,Shirts:PRL:2003,Ytreberg:JCP:2004}

Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules.~\cite{shirts2005comparison,lu2003appropriate}  One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{xx}. Essentially, the method is dominated by contributions of rare snapshots\cite{Jarzynski:PRE:2006,Wu:JCP:2005a,Wu:JCP:2005b}. There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}

@@

\todo[inline]{BA: write section}

@@

\paragraph{Cycle closure error}
Relative free energy calcullations, which compute the change in free energy on making a change to a molecule (e.g. adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking. Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{LOMAP paper from Mobley group; Schrodinger 2015 JACS paper} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways. This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence. Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.

@@

\todo[inline, color={green!20}]{ASJSM:This figure needs to find a better home in the text}
\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}

@@

>>>>

However, this approach ends up being terrible for removing or adding repulsive potentials that go to infinity quickly at or near the origin. One way to look at this is to examine how low $\lambda$ must go to reduce the energy at $0.5\sigma$ (the atomic size parameter) down to 1 $k_BT$, where thermal fluctuations make it possible for other atomic sites to penetrate routinely that deep. If we are trying go from a particle being present, and desire to make it disappear alchemically, then if the repulsive terms is of the form $\epsilon\frac{\sigma}{r}^{12}$, then if $\epsilon$ was 1 $K_BT$ at the temperature of interest, and we start with the particle present, then solving for $(1-\lambda)(1 kB_T)\frac{\sigma}{\frac{1}{2}\sigma}) = 1 k_B T$ we get $\lambda = 1-2^{-12} \sim 0.999976$.  So we have gone virtually all the way to the end of the transformation, there's still an impenetrable post in the middle of our simulation! This is not very much like the desired final state of no interactions between the particle and its environment. We can play around with a few ways of modifying this, like simulating many more intermediate states near $\lambda=1$. Various analyses have shown that this is not a very good strategy, however~\cite{pham2011identifying, beutler1994avoiding, zacharias1994separationshifted, blondel2004ensemble}

@@

The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
$U_0(\vec{q})$ and $U_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
\begin{eqnarray}
\Delta f_{01} & = & \ln \expect{e^{-(U_1(\vec{q}) - u_0(\vec{q}))}}_0 =  \ln \expect{e^{-\Delta u(\vec{q})} }_0
\end{eqnarray}\label{eqn.zwanzig}
and the average is over all samples from the simulation performed with $u_0$. In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is just $U/k_BT$, and $f$ is $A/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
Described in words, we take the samples generated during our run with the potential energy function $U_0(\vec{q})$, recalculate what the energy difference would be if we switched to potential energy function $U_1(\vec{q})$, and average the exponential of the -1 times the energy difference to get -1 times the exponential of the free energy difference.   This almost requires no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $U_1$ in order to calculate the new energies of stored snapshots.  The analysis can be written in a line of code.  We note that this method is even more general, in that the instantaneous work to change the potential energy function from $U_0$ to $U_1$ can be replaced by the non-reversible work $W$ to make the same change under the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, though we do not go into all of the details of nonequilibrum transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble}

Although the Zwanzig equation is formally correct (as long as the two states considered sample the same phase space volume, which is true for standard molecular models), it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules.~\cite{shirts2005comparison,lu2003appropriate}  One can show that if the standard deviation of the difference $\Delta U(\vec{q}) = U_1(vec{q})-U_2(\vec{q})$ over all sampled $\vec{q}$ is large (which in this case, means only several times $k_BT)$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{xx}. Essentially, the method is dominated by contributions of rare snapshots\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}. There are a number of applications where it this expression is nevertheless useful, but those usually involve investigating the dependence of the free energy on small changes in parameters, not changing molecules or even single atoms in molecules~\cite{xx}

@@

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{figures/fig9_convergence/Figure.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}

\paragraph{Cycle closure error}
Relative free energy calculations, which compute the change in free energy on making a change to a molecule (e.g. adding a functional group to a ligand) may provide an additional opportunity for error/consistency checking. Particularly, such calculations are often done to span a graph or tree of free energy calculations~\cite{LOMAP paper from Mobley group; Schrodinger 2015 JACS paper} and in some cases the free energy change to go between molecules A and B can be obtained via multiple transformation pathways. This allows a type of consistency checking where we assess how much the free energy change for that transformation in practice differs from equivalence. Significant deviations from this typically indicate insufficient configurational sampling along the lambda schedule of one or more of the transformations involved.

@@

@@

\todo[inline, color={green!20}]{ASJSM:Write conclusions}
